---
title: 'CSCI E-63C: Week 8 Problem Set'
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
library(cluster)
library(ISLR)
library(MASS)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
options(width = 200)
```

# Preface

In this problem set we will exercise some of the unsupervised learning approaches on [2018 Global Health Observatory (GHO) data](https://www.who.int/gho/publications/world_health_statistics/2018/en/).  It is available at that website in the form of [Excel file](https://www.who.int/gho/publications/world_health_statistics/2018/whs2018_AnnexB.xls?ua=1), but its cleaned up version ready for import into R for further analyses is available at CSCI E-63C canvas course web site [whs2018_AnnexB-subset-wo-NAs.txt](https://canvas.harvard.edu/files/9283256/download?download_frd=1).  The cleaning and reformatting included: merging data from the three parts of Annex B, reducing column headers to one line with short tags, removal of ">", "<" and whitespaces, conversion to numeric format, removal of the attributes with more than 20% of missing values and imputing the remaining missing values to their respective medians.  You are advised to save yourself that trouble and start from preformatted text file available at the course website as shown above.  The explicit mapping of variable names to their full description as provided in the original file is available in Excel file [whs2018_AnnexB-subset-wo-NAs-columns.xls](https://canvas.harvard.edu/files/9283257/download?download_frd=1) also available on the course canvas page.  Lastly, you are advised to download a local copy of this text file to your computer and access it there (as opposed to relying on R ability to establish URL connection to canvas that potentially requires login etc.)

Short example of code shown below illustrates reading this data from a local copy on your computer (assuming it has been copied into current working directory of your R session -- `getwd()` and `setwd()` commands are helpful to find out what is it currently and change it to desired location) and displaying summaries and pairs plot of five (out of almost 40) arbitrary chosen variables.  This is done for illustration purposes only -- the problems in this set expect use of all variables in this dataset.

```{r WHS,fig.height=20,fig.width=20}
whs2018annexBdat <- read.table("whs2018_AnnexB-subset-wo-NAs.txt",sep="\t",header=TRUE,quote="")
summary(whs2018annexBdat[,c(1,4,10)])
summary(whs2018annexBdat[,c(17,26)])
pairs(whs2018annexBdat[,c(1,4,10,17,26)])
```

In a way this dataset is somewhat similar to the `USArrests` dataset extensively used in ISLR labs and exercises -- it collects various continuous statistics characterizing human population across different territories.  It is several folds larger though -- instead of `r nrow(USArrests)` US states and `r ncol(USArrests)` attributes in `USArrests`, world health statistics (WHS) data characterizes `r nrow(whs2018annexBdat)` WHO member states by `r ncol(whs2018annexBdat)` variables.  Have fun!

The following problems are largely modeled after labs and exercises from Chapter 10 ISLR.  If anything presents a challenge, besides asking questions on piazza (that is always a good idea!), you are also encouraged to review corresponding lab sections in ISLR Chapter 10.

# Problem 1: Principal components analysis (PCA) (25 points)

The goal here is to appreciate the impact of scaling of the input variables on the result of the principal components analysis.  To that end, you will first survey means and variances of the attributes in this dataset (sub-problem 1a) and then obtain and explore results of PCA performed on data as is and after centering and scaling each attribute to zero mean and standard deviation of one (sub-problem 1b).

## Sub-problem 1a: means and variances of WHS attributes (5 points)

Compare means and variances of the *untransformed* attributes in the world health statisics dataset -- plot of variance vs. mean is probably the best given the number of attributes in the dataset.  Function `apply` allows to apply desired function (e.g. `mean` or `var` or `sd`) to each row or column in the table.  Do you see all `r ncol(whs2018annexBdat)` attributes in the plot, or at least most of them?  (Remember that you can use `plot(inpX,inpY,log="xy")` to use log-scale on both horizontal and vertical axes.)  Is there a dependency between attributes' averages and variances? What is the range of means and variances when calculated on untransformed data?  Which are the top two attributes with the highest mean or variance?  What are the implications for PCA rendition of this dataset (in two dimensions) if applied to untransformed data?

```{r}
head(whs2018annexBdat)
```

```{r}
attrmeans <- apply(whs2018annexBdat,2,mean)
attrvars <- apply(whs2018annexBdat,2,var)
```

```{r}
plot(attrmeans,attrvars,log="xy")
```

There appears to be a strong postitive linear dependence between the means and variances of the attributes.

```{r}
cat("The min and max values for the means is:",range(attrmeans))
cat("\nThe range for the means is:",abs(range(attrmeans)[1]-range(attrmeans)[2]))
cat("\nThe min and max values for the variances is:",range(attrvars))
cat("\nThe range for the variance is:",abs(range(attrvars)[1]-range(attrvars)[2]))

```

The ranges are quite large therefore scaling is required.

```{r}
#Largest Means
head(sort(attrmeans, decreasing = TRUE))
```

```{r}
#Largest Vars
head(sort(attrvars, decreasing = TRUE))
```

The two largest means and variances are NTDinterventions and TotalPopulation.

The implications of using a PCA in 2D on this dataset are that it will allow us to better see patterns and grouping in the data. By using the two highest variance attributes to plot the points, we are visually removing some of the random noise caused by the highest variance attributes.  

## Sub-problem 1b: PCA on untransformed and scaled WHS data (20 points)

Perform the steps outlined below *both* using *untransformed* data and *scaled* attributes in WHS dataset (remember, you can use R function `prcomp` to run PCA and to scale data you can either use as input to `prcomp` the output of `scale` as applied to the WHS data matrix or call `prcomp` with parameter `scale` set to `TRUE`). To make it explicit, the comparisons outlined below have to be performed first on the unstransformed WHS data and then again on scaled WHS data -- you should obtain two sets of results that you could compare and contrast.

1. Obtain results of principal components analysis of the data (by using `prcomp`)

```{r}
pca.unscaled <- prcomp(whs2018annexBdat,scale=FALSE)
```

```{r}
pca.scaled <- prcomp(whs2018annexBdat,scale=TRUE)
```


2. Generate scree plot of PCA results (by calling `plot` on the result of `prcomp`)

```{r}
plot(pca.unscaled)
```

```{r}
plot(pca.scaled)
```


3. Generate plot of the two first principal components using `biplot`.  Which variables seem to predominantly drive the results of PCA when applied to untransformed data?
  + Please note that in case of untransformed data you should expect `biplot` to generate substantial number of warnings.  Usually in R we should pay attention to these and understand whether they indicate that something went wrong in our analyses.  In this particular case they are expected -- why do you think that is?
  
```{r}
biplot(pca.unscaled, scale = 0)
plot(pca.unscaled$x[,1:2])
text(pca.unscaled$x["United States of America",1],pca.unscaled$x["United States of America",2], "USA")
text(pca.unscaled$x["United Kingdom",1],pca.unscaled$x["United Kingdom",2], "UK")
text(pca.unscaled$x["China",1],pca.unscaled$x["China",2], "China")
text(pca.unscaled$x["India",1],pca.unscaled$x["India",2], "India")
text(pca.unscaled$x["Mexico",1],pca.unscaled$x["Mexico",2], "Mexico")
text(pca.unscaled$x["Australia",1],pca.unscaled$x["Australia",2], "Australia")
text(pca.unscaled$x["Israel",1],pca.unscaled$x["Israel",2], "Israel")
text(pca.unscaled$x["Italy",1],pca.unscaled$x["Italy",2], "Italy")
text(pca.unscaled$x["Ireland",1],pca.unscaled$x["Ireland",2], "Ireland")
text(pca.unscaled$x["Sweden",1],pca.unscaled$x["Sweden",2], "Sweden")
```

The errors were generated because the data is not scaled.

```{r}
biplot(pca.scaled, scale = 0)
plot(pca.scaled$x[,1:2])
text(pca.scaled$x["United States of America",1],pca.scaled$x["United States of America",2], "USA")
text(pca.scaled$x["United Kingdom",1],pca.scaled$x["United Kingdom",2], "UK")
text(pca.scaled$x["China",1],pca.scaled$x["China",2], "China")
text(pca.scaled$x["India",1],pca.scaled$x["India",2], "India")
text(pca.scaled$x["Mexico",1],pca.scaled$x["Mexico",2], "Mexico")
text(pca.scaled$x["Australia",1],pca.scaled$x["Australia",2], "Australia")
text(pca.scaled$x["Israel",1],pca.scaled$x["Israel",2], "Israel")
text(pca.scaled$x["Italy",1],pca.scaled$x["Italy",2], "Italy")
text(pca.scaled$x["Ireland",1],pca.scaled$x["Ireland",2], "Ireland")
text(pca.scaled$x["Sweden",1],pca.scaled$x["Sweden",2], "Sweden")
```

  
4. The field `rotation` in the output of `prcomp` contains *loadings* of the 1st, 2nd, etc. principal components (PCs) -- that can interpreted as contributions of each of the attributes in the input data to each of the PCs.
  + What attributes have the largest (by their absolute value) loadings for the first and second principal component?
  + How does it compare to what you have observed when comparing means and variances of all attributes in the world health statistics dataset?
  
```{r}
cat("The largest loadings for PC1 are:\n")
sort(abs(pca.unscaled$rotation[,1]), decreasing = TRUE)[1:2]

cat("\nThe largest loadings for PC2 are:\n")
sort(abs(pca.unscaled$rotation[,2]), decreasing = TRUE)[1:2]
```

```{r}
pca.unscaled$rotation[,1:2]
```


These are the same two attribuates identified in part 1A have the largest means and variances.

```{r}
cat("The largest loadings for PC1 are:\n")
sort(abs(pca.scaled$rotation[,1]), decreasing = TRUE)[1:2]

cat("\nThe largest loadings for PC2 are:\n")
sort(abs(pca.scaled$rotation[,2]), decreasing = TRUE)[1:2]
```

```{r}
pca.scaled$rotation[,1:2]
```


With the scaled data, I found the two largest PCAs are different from the results in part 1A.  The large absolute size of the unscaled data was masking a large variance on a smaller absolute range.  

5. Calculate percentage of variance explained (PVE) by the first five principal components (PCs).  You can find an example of doing this in ISLR Chapter 10.4 (Lab 1 on PCA).

```{r}
pca.unscaled.var <- pca.unscaled$sdev[1:5]^2
pca.unscaled.PVE <- pca.unscaled.var / sum(pca.unscaled.var)
cat("Unscaled PVE of first five PC:", pca.unscaled.PVE)
cat("\nUnscaled PVE total for first five PC:", round(sum(pca.unscaled.PVE),10))
```

The total PVE essentially becomes 1 for the unscaled data with 99.999% explained by the first PC.

```{r}
pca.scaled.var <- pca.scaled$sdev[1:5]^2
pca.scaled.PVE <- pca.scaled.var / sum(pca.scaled.var)
cat("Scaled PVE of first five PC:", pca.scaled.PVE)
cat("\nScaled PVE total for first five PC:", round(sum(pca.scaled.PVE),10))
```

Now the PVE still becomes 1 with all five PCs however there is a more equitable distribution between the PCs.





Now that you have PCA results when applied to untransformed and scaled WHS data, please comment on how do they compare and what is the effect of scaling?  What dataset attributes contribute the most (by absolute value) to the top two principal components in each case (untransformed and scaled data)?  What are the signs of those contributions?  How do you interpret that?

Please note, that the output of `biplot` with almost 200 text labels on it can be pretty busy and tough to read.  You can achieve better control when plotting PCA results if instead you plot the first two columns of the `x` attribute in the output of `prcomp` -- e.g. `plot(prcomp(USArrests,scale=T)$x[,1:2])`.  Then given this plot you can label a subset of countries on the plot by using `text` function in R to add labels at specified positions on the plot.  Please feel free to choose several countries of your preference and discuss the results.  Alternatively, indicate US, UK, China, India, Mexico, Australia, Israel, Italy, Ireland and Sweden and discuss the results.  Where do the countries you have plotted fall in the graph?  Considering what you found out about contributions of different attributes to the first two PCs, what do their positions tell us about their (dis-)similarities in terms of associated health statistics?





//ANSWER CONTINUED 1B//

The scaling indentified two completely different attributes as PC 1 and PC 2.  The scaling unmasked a larger variance that was covered by the high absolute values of the unscaled data.  The top two attributes that contribute the most by PC are (and their signs):

PC1 Unscaled:
NTDinterventions (positive)    TotalPopulation (positive)

Countries with positive PC1 scores in this model will tend to have a higher number of NTDinteventions and a larger TotalPopulation and vice versa.

PC1 Scaled:
LifeExpectancyF (negative)     HealthyLifeExpectancy (negative)

Countries with negative PC1 scores in this model will tend to have larger values of LifeExpenctancyF (Females) and HealthyLifeExpectancy and vice versa.

PC2 Unscaled:
TotalPopulation (positive)     NTDinterventions (negative)  

Countries with positive PC2 scores in this model will tend to have a higher number of NTDinteventions and a larger TotalPopulation and vice versa.

PC2 Scaled:
CHEperCapita (negative)        CHEpercGDP (negative)

Countries with negative PC1 scores in this model will tend to have larger values of CHEperCapita and CHEperGDP and vice versa.

# Problem 2: K-means clustering (20 points)

The goal of this problem is to practice use of K-means clustering and in the process appreciate the variability of the results due to different random starting assignments of observations to clusters and the effect of parameter `nstart` in alleviating it.

## Sub-problem 2a: k-means clusters of different size (5 points)

Using function `kmeans` perform K-means clustering on *explicitly scaled* (e.g. `kmeans(scale(x),2)`) WHS data for 2, 3 and 4 clusters.  Use `cluster` attribute in the output of `kmeans` to indicate cluster membership by color and/or shape of the corresponding symbols in the plot of the first two principal components generated independently on the same (scaled WHS) data.  E.g. `plot(prcomp(xyz)$x[,1:2],col=kmeans(xyz,4)$cluster)` where `xyz` is input data.  Describe the results.  Which countries are clustered together for each of these choices of $K$?

```{r}
for (i in 2:4){
plot(pca.scaled$x[,1:2],col=kmeans(scale(whs2018annexBdat),i)$cluster)
text(pca.scaled$x["United States of America",1],pca.scaled$x["United States of America",2], "USA")
text(pca.scaled$x["United Kingdom",1],pca.scaled$x["United Kingdom",2], "UK")
text(pca.scaled$x["China",1],pca.scaled$x["China",2], "China")
text(pca.scaled$x["India",1],pca.scaled$x["India",2], "India")
text(pca.scaled$x["Mexico",1],pca.scaled$x["Mexico",2], "Mexico")
text(pca.scaled$x["Australia",1],pca.scaled$x["Australia",2], "Australia")
text(pca.scaled$x["Israel",1],pca.scaled$x["Israel",2], "Israel")
text(pca.scaled$x["Italy",1],pca.scaled$x["Italy",2], "Italy")
text(pca.scaled$x["Ireland",1],pca.scaled$x["Ireland",2], "Ireland")
text(pca.scaled$x["Sweden",1],pca.scaled$x["Sweden",2], "Sweden")
}
```

The more developed nations are generally clustered on the left side of the chart in the green cluster and the more developing nations are to their right.

In the K=2 plot, all the nations other than India are in one cluster.  This does not look correct visually or intuitively.  

In the K=3 plot, the most developed nations are still in one cluster while Mexico and China have broken off into their own cluster.  India is further to the right and also in its own cluster.  

In the K=4 plot, the cluster seem the most accurate.  There is now a cluster separation at the apex of the curve which makes more sense and make the chart more symmetric.  The two blue dots in the center don't seem like they are properly assigned to a cluster however although they have been assigned to India's cluster.  

## Sub-problem 2b: variability of k-means clustering and effect of `nstart` parameter (15 points)

By default, k-means clustering uses random set of centers as initial guesses of cluster centers.  Here we will explore variability of k-means cluster membership across several such initial random guesses.  To make such choices of random centers reproducible, we will use function `set.seed` to reset random number generator (RNG) used in R to make those initial guesses to known/controlled initial state.

Using the approach defined above, repeat k-means clustering of *explicitly scaled* WHS data with four (`centers=4`) clusters three times resetting RNG each time with `set.seed` using seeds of 1, 2 and 3 respectively (and default value of `nstart=1`).  Indicate cluster membership in each of these three trials on the plot of the first two principal components using color and/or shape as described above.  Two fields in the output of `kmeans` -- `tot.withinss` and `betweenss` -- characterize within and between clusters sum-of-squares.  Tighter clustering results are those which have smaller ratio of within to between sum-of-squares.  What are the resulting ratios of within to between sum-of-squares for each of these three k-means clustering results (with random seeds of 1, 2 and 3)?

Please bear in mind that the actual cluster identity is assigned randomly and does not matter -- i.e. if cluster 1 from the first run of `kmeans` (with random seed of 1) and cluster 4 from the run with the random seed of 2 contain the same observations (country/states in case of WHS dataset), they are *the same* clusters.

Repeat the same procedure (k-means with four clusters for RNG seeds of 1, 2 and 3) now using `nstart=100` as a parameter in the call to `kmeans`.  Represent results graphically as before.  How does cluster membership compare between those three runs now?  What is the ratio of within to between sum-of-squares in each of these three cases?  What is the impact of using higher than 1 (default) value of `nstart`?  What is the ISLR recommendation on this offered in Ch. 10.5.1?

One way to achieve everything this sub-problem calls for is to loop over `nstart` values of 1 and 100, for each value of `nstart`, loop over RNG seeds of 1, 2 and 3, for each value of RNG seed, reset RNG, call `kmeans` and plot results for each combination of `nstart` and RNG seed value.

```{r}
for(i in 1:3){

set.seed(i)

plot(pca.scaled$x[,1:2],col=kmeans(scale(whs2018annexBdat),centers=4)$cluster)
text(pca.scaled$x["United States of America",1],pca.scaled$x["United States of America",2], "USA")
text(pca.scaled$x["United Kingdom",1],pca.scaled$x["United Kingdom",2], "UK")
text(pca.scaled$x["China",1],pca.scaled$x["China",2], "China")
text(pca.scaled$x["India",1],pca.scaled$x["India",2], "India")
text(pca.scaled$x["Mexico",1],pca.scaled$x["Mexico",2], "Mexico")
text(pca.scaled$x["Australia",1],pca.scaled$x["Australia",2], "Australia")
text(pca.scaled$x["Israel",1],pca.scaled$x["Israel",2], "Israel")
text(pca.scaled$x["Italy",1],pca.scaled$x["Italy",2], "Italy")
text(pca.scaled$x["Ireland",1],pca.scaled$x["Ireland",2], "Ireland")
text(pca.scaled$x["Sweden",1],pca.scaled$x["Sweden",2], "Sweden")

cat("\nNew Seed")
cat("\ntot.withinss:",kmeans(scale(whs2018annexBdat),centers=4)$tot.withinss)
cat("\nbetweenss:",kmeans(scale(whs2018annexBdat),centers=4)$betweenss)
cat("\nRatio:",(kmeans(scale(whs2018annexBdat),centers=4)$tot.withinss)/kmeans(scale(whs2018annexBdat),centers=4)$betweenss)
cat("\n")

}
```

The iteration with seed=1 has the lowest ratio.  It also visually appears to have the least overlap between clusters and relatively evenly sized clusters.



```{r}
for(i in 1:3){

set.seed(i)

plot(pca.scaled$x[,1:2],col=kmeans(scale(whs2018annexBdat),centers=4,nstart = 100)$cluster)
text(pca.scaled$x["United States of America",1],pca.scaled$x["United States of America",2], "USA")
text(pca.scaled$x["United Kingdom",1],pca.scaled$x["United Kingdom",2], "UK")
text(pca.scaled$x["China",1],pca.scaled$x["China",2], "China")
text(pca.scaled$x["India",1],pca.scaled$x["India",2], "India")
text(pca.scaled$x["Mexico",1],pca.scaled$x["Mexico",2], "Mexico")
text(pca.scaled$x["Australia",1],pca.scaled$x["Australia",2], "Australia")
text(pca.scaled$x["Israel",1],pca.scaled$x["Israel",2], "Israel")
text(pca.scaled$x["Italy",1],pca.scaled$x["Italy",2], "Italy")
text(pca.scaled$x["Ireland",1],pca.scaled$x["Ireland",2], "Ireland")
text(pca.scaled$x["Sweden",1],pca.scaled$x["Sweden",2], "Sweden")

cat("\nNew Seed")
cat("\ntot.withinss:",kmeans(scale(whs2018annexBdat),centers=4)$tot.withinss)
cat("\nbetweenss:",kmeans(scale(whs2018annexBdat),centers=4)$betweenss)
cat("\nRatio:",(kmeans(scale(whs2018annexBdat),centers=4)$tot.withinss)/kmeans(scale(whs2018annexBdat),centers=4)$betweenss)
cat("\n")

}
```

With nstart=100 the lowest ratio now belongs to the seed=3 and the results between the three graphs are very similar.  Their ratios are also closer together.  

The book recommends, "always running K-means cluster with a large value of nstart, such as 20 or 50, since otherwise an undesirable local optimum may be obtained."  

I believe that this was the case earlier and that the current chart is optimal. 

```{r}
for(i in 1:100){
  for (j in 1:3){

set.seed(j)

plot(pca.scaled$x[,1:2],col=kmeans(scale(whs2018annexBdat),centers=4,nstart = i)$cluster)
text(pca.scaled$x["United States of America",1],pca.scaled$x["United States of America",2], "USA")
text(pca.scaled$x["United Kingdom",1],pca.scaled$x["United Kingdom",2], "UK")
text(pca.scaled$x["China",1],pca.scaled$x["China",2], "China")
text(pca.scaled$x["India",1],pca.scaled$x["India",2], "India")
text(pca.scaled$x["Mexico",1],pca.scaled$x["Mexico",2], "Mexico")
text(pca.scaled$x["Australia",1],pca.scaled$x["Australia",2], "Australia")
text(pca.scaled$x["Israel",1],pca.scaled$x["Israel",2], "Israel")
text(pca.scaled$x["Italy",1],pca.scaled$x["Italy",2], "Italy")
text(pca.scaled$x["Ireland",1],pca.scaled$x["Ireland",2], "Ireland")
text(pca.scaled$x["Sweden",1],pca.scaled$x["Sweden",2], "Sweden")

}
}
```

Here we see the higher the nstart value, the more consistent the charts become between seeds.  This indicates that the higher nstart value is truly optimal.  

# Problem 3: Hierarchical clustering (15 points)

## Sub-problem 3a: hierachical clustering by different linkages (10 points)

Cluster country states in (scaled) world health statistics data using default (Euclidean) distance and "complete", "average", "single" and "ward" linkages in the call to `hclust`.  Plot each clustering hierarchy, describe the differences.  For comparison, plot results of clustering *untransformed* WHS data using default parameters (Euclidean distance, "complete" linkage) -- discuss the impact of the scaling on the outcome of hierarchical clustering.

```{r fig.height=15}
plot(hclust(dist(scale(whs2018annexBdat)),method = "complete"))
```

```{r fig.height=15}
plot(hclust(dist(scale(whs2018annexBdat)),method = "average"))
```

```{r fig.height=15}
plot(hclust(dist(whs2018annexBdat),method = "single"))
```

```{r fig.height=15}
plot(hclust(dist(scale(whs2018annexBdat)),method = "ward.D2"))
```

```{r fig.height=15}
plot(hclust(dist(whs2018annexBdat),method = "complete"))
```

As you can see from the above, there is almost no true clusters in the heirarchy for the unscaled data.

## Sub-problem 3b: compare k-means and hierarchical clustering (5 points)

Using function `cutree` on the output of `hclust` determine assignment of the countries in WHS dataset into top four clusters when using Euclidean distance and Ward linkage. (Feel free to choose which one of the two varieties of Ward linkage available in `hclust` you want to use here!).  Use function `table` to compare membership of these clusters to those produced by k-means clustering with four clusters in the Problem 2(b) when using `nstart=100` (and any of the RNG seeds) above.  Discuss the results.

```{r}
cutree(hclust(dist(scale(whs2018annexBdat)),method = "ward.D2"),k=4)
```

```{r}

```


```{r}
table(cutree(hclust(dist(scale(whs2018annexBdat)),method = "ward.D2"),k=4))
table(kmeans(scale(whs2018annexBdat),centers=4,nstart = 100)$cluster)
```

Significantly different results are obtained as seen in the table above.  The group sizes vary significantly although the group numbers do not translate directly.  
